Detecting ditches using machine learning on high-resolution DEMs

### 2020-10-10 LaTeX/anonymised.tex
Completes major revision.

\author{
\name{Anonymised.}}

@@

\begin{abstract}
In this study, a method was developed for detecting ditches in high resolution digital elevation models derived from LiDAR scans.

Thresholding methods using digital terrain indices, such as Sky View Factor, Impoundment Index, or High Pass Median Filter can be used to detect ditches. However, a single threshold generally does not capture the variability in the landscape and generates many false positives and negatives. We hypothesised that, by combining the digital terrain indices using machine learning, we could improve ditch detection at a landscape-scale.

In addition to the raw terrain indices, additional input variables were generated by transforming the data by including neighbouring cells to improve the predictability of the ditches. Random Forests was used to locate the ditches, and the probability output from this classifier was processed to remove noise, and binarised to produce the final ditch prediction.

The confidence interval for the Cohen's Kappa index ranged \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}. Input variables based on the Impoundment Index overall represented the strongest ditch predictors.

The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale.
\end{abstract}

@@

Detection of artificial ditches in a study area with substantial natural stream coverage is
challenging but highly desired for effective forest management. This study developed and applied a methodology using multiple terrain indices and machine learning for automatic ditch detection at the Krycklan catchment in Sweden. The confidence interval for the Cohen's Kappa index ranged [0.655 , 0.781] between the evaluation plots with a confidence level of 95\%. Future work could introduce more input variables and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model, or shape indices to remove stream channels from the prediction. Moving towards image segmentation with grids of pixels instead of pixel classification of tabular pixel values could also be a suitable future approach. With minor adjustments, our proposed method can be employed in a general setting, reducing manual labour and cost, and enabling faster and easier mapping of ditches.

@@

\label{lidartodem}
The LiDAR dataset used in this study (anonymised), and is freely available online at \href{https://gis.krycklan.se/}{https://gis.krycklan.se/}. The average density of the dataset was 20 pulses per $m^2$.  Ground point classification was performed with lasground (included in LAStools, a software suite produced by rapidlasso GmbH). The DEM was created with blast2dem (included in LAStools),  with cell size of $0.5*0.5$ m. Only ground points were imported  to generate a DEM of the ground. This was conducted for 488 las-files of the study area that were mosaicked to one large file using Mosaic to new raster in Arc GIS Pro \citep{EsriArcGisBook}.

>>>>

\author{Anonymised.}

@@

\begin{abstract}
In this study, we develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans. Thresholding methods using digital terrain indices, such as Sky View Factor, Impoundment Index, or High Pass Median Filter can be used to detect ditches. However, a single threshold generally does not capture the variability in the landscape and generates many false positives and negatives. We hypothesise that, by combining the digital terrain indices using machine learning, we can improve ditch detection at a landscape-scale. In addition to the raw terrain indices, additional input variables are generated by transforming the data by including neighbouring cells to improve the predictability of the ditches. Random Forests is used to locate the ditches, and the probability output from this classifier is processed to remove noise, and binarised to produce the final ditch prediction. The confidence interval for the Cohen's Kappa index ranges \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}. Input variables based on the Impoundment Index overall represents the strongest ditch predictors. The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale.
\end{abstract}

@@

%Detection of artificial ditches in a study area with substantial natural stream coverage is challenging but highly desired for effective forest management. This study developed and applied a methodology using multiple terrain indices and machine learning for automatic ditch detection at the Krycklan catchment in Sweden. Despite a substantial vegetation (87\% forest cover), our method achieved good results; the confidence interval for the Cohen's Kappa index ranged [0.655 , 0.781] between the evaluation plots with a confidence level of 95\%. With the developed input variables and extensive post-processing, we managed to remove many stream channels and fill gaps in the ditches where the LiDAR scans were weak. However, these issues are still present to some extent. Future work could introduce more input variables and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model, or shape indices to remove stream channels from the prediction. Moving towards image segmentation with grids of pixels instead of pixel classification of tabular pixel values could also be a suitable future approach. With minor adjustments, our proposed method can be employed in a general setting, reducing manual labour and cost, and enabling faster and easier mapping of ditches.

Detection of artificial ditches over large areas with substantial natural stream coverage is challenging, especially under tree canopy, but highly desired for effective forest management. Ditches represent elongated depressions in a DEM; however, the exact shape and size differs between ditches. We combined recently developed digital terrain indices (the most important being Impoundment Index and High Pass Medium Filter) but, unlike previous studies, did not apply generic thresholding on them. Instead, we included local variability in the indices by using different radii to capture ditches with varied shapes. This integration of a large set of terrain indices using machine learning produced good results, as evidenced by the confidence interval for the Cohen's $\kappa$ index ranging from 0.655 to 0.781 with a confidence level of 95\%. With the developed input variables and extensive post-processing, we also managed to remove many stream channels and fill gaps in the ditches where the LiDAR scans were weak.

Although there is room for further improvement, the methodology is a significant step forward as it performed on a par with previous studies conducted in more homogeneous agricultural landscapes. Our approach enables accurate ditch detection in complex forested landscapes with tree canopy and many natural streams. Future work could introduce more input variables and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model or shape indices to remove stream channels from the prediction. Moving towards image segmentation instead of pixel classification of tabular pixel values could also be a suitable future approach. Our developed methodology can be adjusted to apply to any setting for artificial ditch detection, reducing time and labour investment compared to the conventional manual mapping techniques. The accurate ditch maps can significantly contribute to practical forestry and operational land management at a landscape-scale.

@@

\label{lidartodem}
The LiDAR dataset used in this study was produced by (anonymised), and is freely available online at \href{https://gis.krycklan.se/}{https://gis.krycklan.se/}. The average density of the dataset was 20 pulses per $m^2$.  Ground point classification was performed with lasground (included in LAStools, a software suite produced by rapidlasso GmbH). The DEM was created with blast2dem (included in LAStools),  with cell size of $0.5*0.5$ m. Only ground points were imported  to generate a DEM of the ground. This was conducted for 488 las-files of the study area that were mosaicked to one large file using Mosaic to new raster in Arc GIS Pro \citep{EsriArcGisBook}.
